Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-12 paper.tex
typo

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to better illuminate their ``Section 5'' scene.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{mercadier:PhD}, adding support
for fixslicing is next on the list.

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidences tend to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 5'' scene to exhibit a
bitsliced witness in all the gory details.

@@

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{shiftrows}
    \caption{Shiftrows (from~\citet{beierle:skinny})}
    \label{fig:shiftrows}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{mixcolumns}
    \caption{Mixcolumns (from~\citet{beierle:skinny})}
    \label{fig:mixcolumns}
  \end{subfigure}
\end{figure}

@@

One solution\footnote{Needless to say, in practice, we actually
started off from the simplified implementation, kindly provided
by~\citet{adominicai:fixslicing-gift}, and then used Coq to
understand how these solutions could have been derived.}, which
consists in using Coq as a term rewriting engine, is to postulate the
existence of such an explicit form, let us call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
foldable functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the definition of \coqe!mixcolumns_mod_explicit! from
the proof goal.

>>>>

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to better illuminate their ``Section 4'' scene.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{mercadier:PhD}, adding support
for fixslicing is next on the list.

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidences tend to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 4'' scene to exhibit a
bitsliced witness in all the gory details.

@@

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{shiftrows}
    \caption{Shiftrows (from~\citet{beierle:skinny})}
    \label{fig:shiftrows}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{mixcolumns}
    \caption{Mixcolumns (from~\citet{beierle:skinny})}
    \label{fig:mixcolumns}
  \end{subfigure}
  \caption{\Skinny{} diffusion layer}
\end{figure}

@@

One solution\footnote{Needless to say, in reality, we first went
looking find 4 simplified implementations, inspired
by~\citet{adominicai:fixslicing-gift}. Only later did we use Coq to
understand how these solutions \emph{could} have been derived from
first principles.}, which consists in using Coq as a term rewriting
engine, is to postulate the existence of such an explicit form, let us
call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
foldable functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the definition of \coqe!mixcolumns_mod_explicit! from
the proof goal.
